---
title: "Clinical Classifications Software"
author: "Hunter Ratliff, htratlif@utmb.edu"
date: "5/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(tidyverse)
library(icd) # helps with ICD codes
library(touch)
```

# ICD-10

## Procedures - CCS

ICD-10-PCS (procedure) codes --> Clinical Classifications Software

Downloaded _beta_ version 2020.1 from [HCUP website](https://www.hcup-us.ahrq.gov/toolssoftware/ccs10/ccs10.jsp) on 2020-05-20

The beta versions of the ICD-10-PCS tool file (CSV) are composed of eight fields separated by commas. The files are organized as follows:
- Field 1: Procedure Code
- Field 2: CCS Category
- Field 3: Code Description
- Field 4: CCS Category Description
- Field 5: Multi-level 1 Category
- Field 6: Multi-level 1 Category Description
- Field 7: Multi-level 2 Category
- Field 8: Multi-level 2 Category Description

```{r, eval=F}
ccs_pr_icd10 <- read_csv("HCUP Files/ccs_pr_icd10pcs_2020_1.csv",skip = 1,
                         col_names = c("I10_PR", "CCS_category", "I10_PR_Desc", "CCS_category_Desc", 
                                       "CCS_Lvl_1", "CCS_Lvl_1_Label", "CCS_Lvl_2", "CCS_Lvl_2_Label")) %>%
  mutate_all(~str_remove_all(.,"'"))

ccs_pr_icd10 %>% write_csv("HCUP Files/clean_ccs/ccs_pr_icd10.csv")
ccs_pr_icd10


```

## Diagnosis - CCS

This has been replaced by the CCSR^[Clinical Classifications Software **Refined**] (below), and is [archived here](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/ccsr_archive.jsp#ccsbeta). I downloaded the CCS v2019.1 (valid for ICD-10-CM codes through September 2019) on 2020-05-20

Unlike the CCSR, this table maps ICD-10 codes to the same 283 CCS categories as the ICD-9 CCS

```{r, eval=F}
ccs_dx_icd10 <- read_csv("HCUP Files/ccs_dx_icd10cm_2019_1.csv", skip = 1,
                            col_names = c("I10_DX", "CCS_category", "I10_DX_Desc", "CCS_category_Desc", 
                                          "CCS_Lvl_1", "CCS_Lvl_1_Label", "CCS_Lvl_2", "CCS_Lvl_2_Label")) %>%
  mutate_all(~str_remove_all(.,"'"))

ccs_dx_icd10
```


## Diagnosis - CCSR

ICD-10-CM (diagnosis) codes -->  Clinical Classifications Software **Refined**

Downloaded CCSR version 2020.2 from [HCUP website](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/ccs_refined.jsp) on 2020-05-20

These codes are unique from the CCS codes in a few ways, as outlined in [table 1](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/DXCCSR-User-Guide.pdf#page=7) of the documentation. Unlike the CCS codes, these map ICD codes to _multiple_ categories. These are listed in `CCSR_cat(1-5)`, and they've also assigned a default category (that I call `CCSR_default`).

```{r, eval=F}
ccsR_dx_icd10 <- read_csv("HCUP Files/$DXCCSR_v2020-2.CSV", skip=1,
                          col_types = cols(.default = col_character()), 
                          col_names = c("I10_DX", "I10_DX_Desc", "CCSR_default", "CCSR_default_label",
                                        "CCSR_cat1", "CCSR_cat1_label", "CCSR_cat2", "CCSR_cat2_label",
                                        "CCSR_cat3", "CCSR_cat3_label", "CCSR_cat4", "CCSR_cat4_label",
                                        "CCSR_cat5", "CCSR_cat5_label")) %>%
  mutate_all(~str_remove_all(.,"'"))

ccsR_dx_icd10
```

# ICD-9

## Procedures - CCS

```{r}
"TBD"
```


## Diagnosis - CCS


# CPT to CSS

Used to map CPT and HCPCS codes into CCS categories

Downloaded 2019 version from [HCUP website](https://www.hcup-us.ahrq.gov/toolssoftware/ccs_svcsproc/ccssvcproc.jsp) on 2020-05-20. Note that CPT codes are copyright of the American Medical Association. All rights reserved.

It's also a pain since they give you ranges

```{r}
ccs_pr_cpt <- read_csv("HCUP Files/2019_ccs_services_procedures.csv", 
    skip = 2, col_names = c("CodeRange", "CSS", "CSS_Label")) %>%
  mutate_all(~str_remove_all(.,"'")) %>%
  separate(CodeRange, c("CodeRange_begin", "CodeRange_end"))

ccs_pr_cpt
```

It's also a pain since they give you ranges, which are hard to look up with CPT codes (since they aren't all numbers). I made a quick and dirty function below that should get the job done, but didn't test it every much

```{r, warning=F}
cpt_to_css <- function(cpt_code, data) {
  # If data not supplied, use ccs_pr_cpt df
  if(missing(data) & exists("ccs_pr_cpt")) data <- ccs_pr_cpt
  
  # If the code is numeric (or should be but wasn't passed
  # as an interger), make it one
  if(is.numeric(cpt_code)|!str_detect(cpt_code, "[:alpha:]")) {
    cpt_code <- as.numeric(cpt_code)
  }
  
  # If coerced cpt_code is numeric, use ranges
  if(is.numeric(cpt_code)) {
    ccs_pr_cpt %>% 
      mutate(CodeRange_begin = as.numeric(CodeRange_begin),
             CodeRange_end = as.numeric(CodeRange_end)) %>%
      filter(!is.na(CodeRange_begin)) %>%
      filter(cpt_code>=CodeRange_begin, cpt_code<=CodeRange_end)
  } else({
    filter(ccs_pr_cpt, CodeRange_begin==cpt_code  |  CodeRange_end==cpt_code)
  })
}


cpt_to_css(28800)   
cpt_to_css("0202T")  # Results when code isn't numeric
cpt_to_css("27603")  # Still works if numeric code passed as char
cpt_to_css(97597)

rm(cpt_to_css)
```





