---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hcup- Implementation of HCUP Software Tools in R

<!-- badges: start -->
<!-- badges: end -->

Adapts the software provided by the [Healthcare Cost and Utilization Project (HCUP)](https://www.hcup-us.ahrq.gov/tools_software.jsp) 
for analyses of administrative database using International Classification of Diseases 
(ICD) codes within the R statistical environment.

## Installation

You can install the development version of hcup from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("HunterRatliff1/hcup")
```

## Classify ICD-10 codes to CCSR categories

The simple case is for procedures, as each procedure has one CCSR category

```{r CCSR-procedures}
library(hcup)
classify_ccsr_pr("0016070")

icd_10_pr <- c("04L04DZ", "009Y40Z", "0F7C7DZ", "GZ2ZZZZ")
classify_ccsr_pr(icd_10_pr)

# Vectorized using `mutate()`
library(dplyr)

dplyr::tibble(icd_10 = icd_10_pr) %>%
  mutate(CCSR = classify_ccsr_pr(icd_10),
         explained = explain_ccsr(CCSR))
```

CCSR diagnosis codes...
