---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hcup- Implementation of HCUP Software Tools in R

<!-- badges: start -->
<!-- badges: end -->

Adapts the software provided by the [Healthcare Cost and Utilization Project (HCUP)](https://www.hcup-us.ahrq.gov/tools_software.jsp) 
for analyses of administrative database using International Classification of Diseases 
(ICD) codes within the R statistical environment.

## Installation

```{r setup, include=FALSE}
# Hide the warning messages
library(hcup)
library(dplyr)
```


You can install the development version of hcup from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("HunterRatliff1/hcup")
```

## Procedures with ICD-10

Although I'm not a fan of procedures myself (clinical practice), I am aficionado of ICD-10
procedure codes (ICD-10-PCS). Compared to their predecessor (ICD-9), ICD-10-PCS is much more
eloquent and specific, so it's only appropriate that we start there!

### Procedure classification

The most straightforward of the HCUP software tools included in this package are the 
[Procedure Classes](https://www.hcup-us.ahrq.gov/toolssoftware/procedureicd10/procedure_icd10.jsp).
Procedure classes identify procedures as [1] diagnostic or therapeutic and [2] whether they major 
(expected to be performed in an operating room) or minor.

```{r}
library(hcup)
classify_pr("0PS443Z") # Spine surgery
classify_pr("009U3ZX") # Lumbar puncture
```

This function can be vectorized (e.g. using `dplyr::mutate`) as shown below:

```{r}
# Using the dplyr workflow
library(dplyr)
tibble(
  Patients = c("A",       "B",       "C"),
  ICD_10   = c("0PS443Z", "3E05305", "F13ZHZZ")) %>%
  mutate(pr_class = classify_pr(ICD_10))
```

Although applications using ICD-9 codes aren't covered here, this function also 
works the same for ICD-9 procedure codes.

```{r}
classify_pr("0015") # Infusion of IL-2
```

### CCSR for ICD-10-PCS Procedures

Much more powerful is the [Clinical Classifications Software Refined (CCSR)](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/prccsr.jsp) categories, which 
aggregates the expansive number of ICD-10-PCS procedure codes into a smaller number of 
clinically meaningful categories.

Similar to the procedure classes (and all functions starting with `classify_`), this categorizes ICD
codes into mutually exclusive categories. 

```{r}
classify_ccsr_pr("0016070")
```

Again, this can be vectorized for use within data frames. The helper function `explain_ccsr` can
convert these categories into human-readable labels

```{r}
library(dplyr)
tibble(icd_10 = c("04L04DZ", "009Y40Z", "0F7C7DZ", "GZ2ZZZZ")) %>%
  mutate(CCSR      = classify_ccsr_pr(icd_10),
         explained = explain_ccsr(CCSR))
```

You can also use the "detailed" option within `explain_ccsr(ccsr, detailed=TRUE)` to return 
additional data about the CCSR category. This returns a named list, which can be helpful
in conjunction with the [purrr package](https://purrr.tidyverse.org/).

```{r}
library(purrr)
# NOTE: The results returned when `detailed = TRUE` depends on the
#       type of CCSR category. See the `CCSR_PR_categories` or 
#       `CCSR_DX_categories` in the `hcup.data` package for details
tibble(CCSR_PR = c("URN001", "RES010", "CAR004")) %>%
  mutate(Explained = explain_ccsr(CCSR_PR),
         details   = explain_ccsr(CCSR_PR, detailed = TRUE),
         domain    = map_chr(details, "clinical_domain"))
```

