---
title: "CCSR for ICD-10-CM Diagnosis Codes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(hcup)
library(dplyr)
library(tibble)
library(purrr)
library(tidyr)
```

The [Clinical Classifications Software Refined (CCSR) software](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/ccsr_archive.jsp#ccsr) for ICD-10-CM 
diagnosis codes provides both one-to-one mapping (`classify_ccsr_dx1()`) and one-to-many mapping 
(`ccsr_dx()`) to CCSR categories. The one-to-many mapping is necessary because many ICD-10 codes 
span multiple meaningful categories, and the identification of all conditions related to a 
diagnosis code would be lost by categorizing some ICD-10-CM codes into a single category.

# A simple demonstration

For example, consider the code **I11.0** (_Hypertensive heart disease with heart failure_) which
encompasses both **heart failure** (CCSR category CIR019) and **hypertension with complications** 
(CIR008). Classifying this code as heart failure alone would miss meaningful information, but on 
the other hand, some analyses require mutually exclusive categorization.

The `ccsr_dx()` function identifies all CCSR categories associated with an ICD-10 diagnosis code. 
Each ICD-10 code can have one or more CCSR category. In the example used above, I11.0 has two CSSR
categories, so the function returns a character vector of two CCSR categories

```{r setup}
library(hcup)
library(dplyr)
ccsr_dx("I110")
ccsr_dx("I110") %>% explain_ccsr()
```

If we need to do one-to-one mapping (e.g. descriptive statistics on the number of hospital discharges), we use `classify_ccsr_dx1()`. Importantly, this algorithm is only valid for the principal diagnosis (for inpatient
records) or the first listed diagnosis code (for outpatient data)

```{r}
classify_ccsr_dx1("I110", setting = "inpatient")
```

# One-to-One with classify_ccsr_dx1

The `classify_ccsr_dx1()` function is similar to the other `classify_` functions used in this
package, so we will begin here. 

Although a comprehensive review of the CCSR methodology is beyond the scope of this article 
(I'd highly recommend reading the appendices of the [CCSR DX user guide](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/DXCCSR-User-Guide-v2022-1.pdf)), I do 
want to highlight a few important points here.

## Only valid for the principal/first diagnosis

This algorithm was developed for use with the principal diagnosis (for inpatient records) or 
the first listed diagnosis code (for outpatient data). Consider the sample of two patients
below, with their principal diagnosis listed in `DX1`, and subsequent diagnoses listed in 
`DX_n`.

```{r}
library(dplyr)
df <- tibble::tribble(
  ~Pt_id,    ~DX_1,    ~DX_2,  ~DX_3,
     "A",    "I10", "F17210",     NA,
     "B",    "D62",   "Z370", "E876"
  )

## GOOD
df %>%
  mutate(prin_CCSR = classify_ccsr_dx1(DX_1, setting="ip"))

## BAD
df %>%
  mutate(prin_CCSR = classify_ccsr_dx1(DX_2, setting="ip"))
```

Caution should be used when working with tidy data (long-format) to avoid classifying the
non-principal diagnoses

```{r}
df_tidy <- df %>%
  pivot_longer(cols      = -Pt_id, 
               names_to  = "DX_NUM", 
               values_to = "ICD")

## GOOD
df_tidy %>%
  filter(DX_NUM=="DX_1") %>%
  mutate(prin_CCSR = classify_ccsr_dx1(ICD, setting="ip"),
         expln     = explain_ccsr(prin_CCSR))

## BAD
df_tidy %>%
  mutate(prin_CCSR = classify_ccsr_dx1(ICD, setting="ip"),
         expln     = explain_ccsr(prin_CCSR))

rm(df, df_tidy)
```


## Unacceptable principal/first diagnosis

Second, not all ICD-10 codes are acceptable for default classification. For example if **B95.2** 
were listed as the Principal/First diagnoses, it would map to `XXX000` (for inpatient data) 
and `XXX111` (for outpatient data).

```{r}
# B95.2: Enterococcus as the cause of diseases classified elsewhere
classify_ccsr_dx1("B952", setting = "inpatient")
classify_ccsr_dx1("B952", setting = "outpatient")
```

This could be because B95.2 was inappropriately listed as the principal diagnosis, and if
you run into this problem frequently, it may be helpful to consult the [CCSR DX user guide](https://www.hcup-us.ahrq.gov/toolssoftware/ccsr/DXCCSR-User-Guide-v2022-1.pdf).

## Setting isn't that important

As of this version of the CCSR (see `?hcup.data::CCSR_DX_mapping` to check the version),
the setting (i.e. inpatient or outpatient) doesn't drastically change the results. Using 
the dataset this function is based off of, we can see that the only situations where
setting = "inpatient" and setting = "outpatient" don't agree are codes that are acceptable
for the first diagnosis in the outpatient setting, not for the principal diagnosis in the 
inpatient setting.

```{r}
if(interactive()){
  hcup.data::CCSR_DX_mapping %>%
    
    # Recode the Unacceptable category to be the same for ip/op
    mutate(default_CCSR_IP = recode(default_CCSR_IP, "XXX000"="Unccpt"),
           default_CCSR_OP = recode(default_CCSR_OP, "XXX111"="Unccpt")) %>%
    
    # Remove default categories that match across settings
    filter(default_CCSR_IP!=default_CCSR_OP) %>%
    
    count(default_CCSR_IP, default_CCSR_OP, sort=T)
}
```


# One-to-Many with ccsr_dx()

`ccsr_dx()` is different than most of the other functions in this package, because each
ICD code can return multiple categories per ICD code. This means that we'll be dealing 
with lists. To demonstrate, consider the two ICD-10 codes below:

|pt_id    | DX_NUM|`I10_DX` |
|:--------|:------|:--------|
|John Doe |      1|`A401`   |
|John Doe |      2|`K432`   |

The code `A401` maps to two CCSR categories (_INF002_ & _INF003_) while `K432` only has 
one associated CCSR category (_DIG010_)

```{r}
ccsr_dx("A401")
ccsr_dx("K432")
```

Appending this to the previous table gives us

|pt_id    | DX_NUM|`I10_DX` |CCSR           |
|:--------|------:|:--------|:--------------|
|John Doe |      1|`A401`   |INF002, INF003 |
|John Doe |      2|`K432`   |DIG010         |

```{r}
df.listcol <- dplyr::tibble(pt_id = "John Doe",
                            DX_NUM = 1:2,
                            I10_DX = c("A401", "K432")) %>%
  mutate(CCSR = ccsr_dx(I10_DX))

df.listcol

str(df.listcol$CCSR)
```

The column **CCSR** in the table above is now a list. For those unfamiliar with list
columns the book [Functional Programming](https://dcl-prog.stanford.edu/list-columns.html)
has a helpful overview chapter and the [Rectangle vignette](https://tidyr.tidyverse.org/articles/rectangle.html) in tidyr. You can also use `View(df.listcol)` in RStudio to see the
contents of the list column in a more familiar structure.

If you want to get this list-column into an atomic vector, `tidyr::unnest_longer()` will
make a new row for every element within the list

```{r}
df.listcol %>%
  unnest_longer(CCSR)
```

# A longer example

```{r}
df <- tibble::tribble(
  ~pt_id, ~DX_NUM, ~ICD10,
  "A",     1,      "K432",
  "A",     2,      "A401",
  "B",     1,      "I495",
  "B",     2,   "BadCode",
  "C",     1,     "E8771",
  "C",     2,     "A5442",
  "C",     3,      "A564"
)
```

```{r}
df %>%
 mutate(CCSR = ccsr_dx(ICD10)) %>%
 tidyr::unnest_longer(CCSR) %>%
 mutate(CCSR_expl = explain_ccsr(CCSR))
```

